---
title: "Business Forecasting Mid-Term Exam"
author: "Blake Tarasov"
date: "`r Sys.Date()`"
output: html_document
---

# **Introduction**
Public transit ridership is an important metric for evaluating the effectiveness of public transportation systems. This report analyzes the transit ridership dataset from January 2000 using various time series forecasting methods.

## **Import Data**
```{r}
library(readr)
library(ggplot2)
library(forecast)
library(TTR)
library(tseries)

# Read the dataset
TRANSIT <- read_csv("Downloads/TRANSIT.csv")
Transit_Raw <- TRANSIT$TRANSIT
Transit_ts <- ts(Transit_Raw, frequency = 12, start = c(2000,1))

# Plot Time Series
plot(Transit_ts, main="Time Series Plot of Transit Ridership", ylab="Ridership", xlab="Year")
```

## **Plot and Inference**
### **ACF Plot**
```{r}
acf(Transit_ts, main="ACF Plot of Transit Ridership")
```
### **Observations:**
- The time series shows an overall trend with seasonal fluctuations.
- The ACF plot indicates strong autocorrelation, confirming seasonality and trend.

## **Central Tendency**
```{r}
fivenum(Transit_ts) # Min, Q1, Median, Q3, Max
mean(Transit_ts)

# Boxplot
boxplot(Transit_ts, main="Boxplot of Transit Ridership", ylab="Ridership")
```
### **Observations:**
- The distribution is slightly skewed, with some lower outliers around 2020.

## **Decomposition**
```{r}
decomposed <- decompose(Transit_ts)
plot(decomposed)
```
### **Observations:**
- The time series exhibits seasonality.
- The decomposition is additive.
- The highest ridership is in **October**, and the lowest is in **February**.

## **Seasonally Adjusted Time Series**
```{r}
seasonAdjusted <- Transit_ts - decomposed$seasonal
plot(Transit_ts, main="Seasonally Adjusted vs. Actual Time Series")
lines(seasonAdjusted, col="red")
```
### **Observations:**
- Seasonality has a significant effect on transit ridership.

## **Naïve Method Forecasting**
```{r}
Naive <- naive(Transit_ts)
plot(Naive)
```
### **Residual Analysis**
```{r}
# Plot Residuals
residuals_naive <- residuals(Naive)
plot(residuals_naive, main="Residuals Plot", ylab="Residuals", xlab="Time")

# Histogram of Residuals
hist(residuals_naive, main="Histogram of Residuals", xlab="Residuals", col="gray")

# Residuals vs. Fitted Values
fitted_naive <- fitted(Naive)
plot(fitted_naive, residuals_naive, main="Fitted Values vs. Residuals")

# Actual vs. Residuals
plot(Transit_ts, residuals_naive, main="Actual Values vs. Residuals")

# ACF of Residuals
acf(na.omit(residuals_naive), main="ACF of Residuals")
```
### **Accuracy Measures**
```{r}
accuracy(Naive)
```
### **Forecast for Next Year**
```{r}
forecast_naive <- forecast(Transit_ts, h=12)
plot(forecast_naive)
print(forecast_naive)
```

## **Simple Moving Averages**
```{r}
MA3 <- ma(Transit_ts, order = 3)
MA6 <- ma(Transit_ts, order = 6)
MA9 <- ma(Transit_ts, order = 9)

plot(Transit_ts, main="Moving Averages of Transit Ridership")
lines(MA3, col='red')
lines(MA6, col='blue')
lines(MA9, col='green')
```
### **Observations:**
- Higher MA orders smooth the data but introduce lag.

## **Bonus: Forecast Using MA(6)**
```{r}
futureMA6 <- ts(rep(tail(MA6, 1), 12), start = c(2025, 1), frequency = 12)
plot(Transit_ts, xlim=c(2000,2026))
lines(MA6, col='blue')
lines(futureMA6, col='purple', lwd=3, lty=2)
```
### **Final Observations:**
- The forecast suggests a continuation of the post-2020 recovery trend.
- The MA(6) forecast provides a reasonable approximation but does not capture shocks.

---